Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-18 manuscript/paper.tex
structrue and cages figure. and its caption

\documentclass[a4paper,prb,twocolumn]{revtex4-1}  %,twocolumn
%\documentclass[preprint]{revtex4-1}

@@

@@

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of the crystal field Hamiltonian of this system 
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed the same~\cite{paxton-notes}.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 relevant for the d-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$)
 and called octahedral field. 
It is given by~\cite{pavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

%\subsection{cooperative JT and orbital order}
%\az{just a brief description... cooperative JT and orbital order}
To complete the revision of the JT instability,
it is worth mentioning that
JT effect occurs in all kinds of systems, molecules, clusters and crystals.
In crystals, e.g., perovskites, it becomes a cooperative effect where the
elongation of one octahedron is accompanied by a compression of adjacent octahedron,
which minimises the changes in the unit cell volume and hence elastic energy cost of the transformation.
Such alternate elongations and compressions also lead to 
an interesting long range orbital ordering where alternate orbitals
are populated on the adjacent JT sites.

@@

We consider a
$\sqrt{2}\times\sqrt{2}\times 2$ unit cell of the cubic structure that transforms
to the unit cell of the orthorhombic structure under octahedral rotation and tilting.
Assuming $\vec a_1,\vec a_2,\vec a_3$ to be the lattice vectors,
two A and two B atoms in a layer are situated at
$(\vec a_1 + \vec a_3/2)/2$, $(\vec a_2+ \vec a_3/2)/2$,
and $(0,0,0)$, $(\vec a_1 + \vec a_2)/2$, 
while similar atoms in the second layer are shifted by $\vec a_3/2$ from these.
While the lattice vectors change under the rotation and tilting of the octahedra, 
these relations do not change.
%The B atom at the origin will be considered as the central JT atom.
The O atoms in the cubic structure lie half way between the B atoms and make octahedral cages around them.
Their positions in the deformed structure %(see appendix) 
are not directly required for our calculations and discussion.

We rotate the octahedra by an angle $\beta$ about $\hat z \equiv 001$
and then tilt it by rotating it about new $110$ by an angle $\alpha$.
Four octahedra have to be rotated and tilted clockwise or counter closckwise
consistently. For example, %$\alpha  \{1,-1,-1,1\}$, $\beta  \{1,-1,1,-1\}$.
$(\alpha,\beta)$,
$(-\alpha,-\beta)$,
$(-\alpha,\beta)$,
$(\alpha,-\beta)$
for the atoms labelled $1$ through $4$ in Fig.~\ref{fig:structure}.
The structure transforms from cubic to orthorhombic under these rotation and tilting of the octahedra.
The rotation matrix $R$ for $(\alpha,\beta)$ case is given in the appendix~\ref{sec:appendix}. 
The lattice vectors of the deformed structure and the positions of A and B atoms can be calculated easily by calculating the coordinates of all six oxygen atoms that make the octahedra (by rotating and tilting the octahedra) for all four octahedra in the unit cell and 
comparing the coordinates of their shared oxygen atoms.
We obtain,
%\az{just rescaled, not rotated. write as cos * (1,1), etc?}
% lattice vectors:
\begin{align}
\left(
\begin{array}{c}
\vec a_1 \\
\vec a_2\\
\vec a_3\\
\end{array}
\right)
= a
\left(
\begin{array}{ccc}
 \cos \beta  & \cos \beta  & 0 \\
 -\cos \alpha  \cos \beta  & \cos \alpha  \cos \beta  & 0 \\
 0 & 0 & 2 \cos \alpha  \\
\end{array}
\right)
\end{align}

@@

>>>>

\documentclass[a4paper,prb,twocolumn]{revtex4-1}  %a4paper,prb,twocolumn
%\documentclass[preprint]{revtex4-1}

%\documentclass[12pt]{report}

@@

%\usepackage[paperheight=18cm,paperwidth=14cm,textwidth=12cm]{geometry}
%\usepackage{setspace}
%\doublespacing

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-1}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-z}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-z}
%\includegraphics[height=6.cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/AB-cages-deformed}
\caption{
Crystal structures and coordination cages.
%Crystal structure of cubic and orthorhombic perovskites (a-b) and ion coordination cages around B ions (c-d).
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:cages}
\end{center}
\end{figure}

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of its crystal field Hamiltonian
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed the same~\cite{paxton-notes}.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the d-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$)
 and is called octahedral field. 
It is given by~\cite{pavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

%\subsection{cooperative JT and orbital order}
%\az{just a brief description... cooperative JT and orbital order}
To complete the revision of the JT instability,
it is worth mentioning that
JT effect occurs in all kinds of systems---molecules, clusters and crystals.
In crystals, e.g., perovskites, it becomes a cooperative effect where the
elongation of one octahedron is accompanied by a compression of adjacent octahedron,
which minimises the changes in the unit cell volume and hence the elastic energy cost of the transformation.
Such alternate elongations and compressions also lead to 
an interesting long range orbital ordering where alternate orbitals
are populated on the adjacent JT sites.

@@

We consider a
$\sqrt{2}\times\sqrt{2}\times 2$ unit cell of the cubic structure that transforms
to the unit cell of the orthorhombic structure under octahedral rotation and tilting.
Assuming 
%$\vec a_1,\vec a_2,\vec a_3$
$\textbf{a},\textbf{b},\textbf{c}$
 to be the lattice vectors,
two A and two B atoms in a layer are situated at
%$(\vec a_1 + \vec a_3/2)/2$, $(\vec a_2+ \vec a_3/2)/2$,
$(\textbf{a} + \textbf{c}/2)/2$, $(\textbf{b}+ \textbf{c}/2)/2$,
and $(0,0,0)$, 
%$(\vec a_1 + \vec a_2)/2$, 
$(\textbf{a}  + \textbf{b} )/2$, 
while similar atoms in the second layer are shifted by 
%$\vec a_3/2$
$\textbf{c} /2$
 from these.
While the lattice vectors change under the rotation and tilting of the octahedra, 
these relations do not change.
%The B atom at the origin will be considered as the central JT atom.
The O atoms in the cubic structure lie half way between the B atoms and make octahedral cages around them.
Their positions in the deformed structure %(see appendix) 
are not directly required for our calculations and discussion.

We rotate the octahedra by an angle $\beta$ about $\hat z \equiv 001$
and then tilt it by rotating it about new $110$ by an angle $\alpha$.
Four octahedra have to be rotated and tilted clockwise or counter closckwise
consistently. For example, %$\alpha  \{1,-1,-1,1\}$, $\beta  \{1,-1,1,-1\}$.
$(\alpha,\beta)$,
$(-\alpha,-\beta)$,
$(-\alpha,\beta)$,
$(\alpha,-\beta)$
for the atoms labelled $1$ through $4$ in Fig.~\ref{fig:structure}.
The structure transforms from cubic to orthorhombic under these rotation and tilting of the octahedra.
The rotation matrix $R$ for $(\alpha,\beta)$ case is given in the appendix~\ref{sec:appendix}. 
The lattice vectors of the deformed structure and the positions of A and B atoms can be calculated easily by calculating the coordinates of all six oxygen atoms that make the octahedra (by rotating and tilting the octahedra) for all four octahedra in the unit cell and 
comparing the coordinates of their shared oxygen atoms.
We obtain,
%\az{just rescaled, not rotated. write as cos * (1,1), etc?}
% lattice vectors:
\begin{align}
\left(
\begin{array}{c}
%\vec a_1 \\
%\vec a_2\\
%\vec a_3\\
\textbf{a}\\
\textbf{b}\\
\textbf{c}
\end{array}
\right)
= a
\left(
\begin{array}{ccc}
 \cos \beta  & \cos \beta  & 0 \\
 -\cos \alpha  \cos \beta  & \cos \alpha  \cos \beta  & 0 \\
 0 & 0 & 2 \cos \alpha  \\
\end{array}
\right)
\end{align}

@@

\az{V20, JT:\\
V2-2, V4-2, couple yz and zx.
But... rotation.... produce all m components in general, ....
}
